---
title: "Neural Networks practical"
author: "Marc Sabaté"
date: "08/03/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Deep Learning practical in R
In this practical we will build and train a feedforward network with sigmoid activated neurons from scratch.


### Load the necessary libraries
We load ggplot2 for plotting purposes. If `ggplot2` is not installed, run first `install.packages('ggplot2')`.
```{r}
library(ggplot2) 
```

### Data loading and visualisation
Create a directory *DL_practical* in your home directory and save the file *data.txt*.
We read the data:
```{r}
HOME = Sys.getenv(x='HOME')
setwd(paste(HOME,"DL_practical",sep="/"))
data = read.table("data.txt", sep=",", header = TRUE)
```

We explore the data. As you can see, each observation of the dataset is two-dimensional with a binary output. 
The goal of this practical is to fit a binary classifier on the data.
```{r}
summary(data)
table(data$y)
```

Execute the following lines to visualise it. 
```{r}
p = ggplot(data=data, aes(x1,x2, color=factor(y))) + geom_point() 
print(p)
```

### Logistic Regression
As a first choice, we try to fit a Logistic regression on the data. For this, we use the R function `glm()` taking into account that its parameter `family` providing the error distribution should be `'binomial'`.
```{r}
model = glm(formula = y~x1+x2, data = data, family = "binomial")
```

We create a grid in order to test our model. The output of the Logistic regression is a numeric value between 0 and 1, providing the probability of the observation being 1. In other words, $f_{\theta}(x_1,x_2) = P(Y=1|x_1,x_2;\theta)$.
Therefore, we create the binary variable `y_grid_binary`, that outputs the class `1` when the probability is greater than 0.5.
```{r}
grid = expand.grid(x1 = seq(min(data$x1), max(data$x1), 0.02),
                   x2 = seq(min(data$x2), max(data$x2), 0.02))

y_grid = predict(model, grid, type="response")
y_grid_binary = as.numeric(y_grid>0.5)
grid$y = y_grid_binary
```

We plot the prediction of the model, along with the training set, in order to see how well the logistic regression is doing. 
```{r}
p = ggplot(data=data, aes(x1,x2, color=factor(y))) + geom_point(size=1) 
p = p + geom_point(data=grid, aes(x1, x2, color=factor(y)), alpha=0.1)
print(p)
```

As you can see, the Logistic Regression is trying to linearly separate the training set, and therefore it doesn't work well in those datasets that are not linearly separable. 

### Neural Networks
#### Ingredients
- Definition of the model to do a forward pass: $y = \sigma(W^{[L-1]}(\sigma(W^{[L-2]}(... \sigma(W^{[1]}x + b^{[1]})...)+b^{[L-2]}))+b^{[L-1]})$ where 
    - $\sigma$ is the activation function
    - $Wx + b$ is a linear function that goes from $\mathbb{R}^n$ to $\mathbb{R}^m$.
    - In our example, the output of the model is a value between 0 and 1 that tells the probability of a point being blue or red. The input of the model are the coordinates of the point. 
- Definition of a cost function that tells how good is the model in terms of its parameters: 

$$J(\theta), \text{where }\theta:=\{W^{[1]}, b^{[1]}, \ldots, W^{[L-1]}, b^{[L-1]}\}$$

- Optimisation of the cost function using gradient descent.
    - We need to apply the chain rule (backpropagation) in order to obtain $\partial_{\theta}J$ for each optimisation step.
    
#### Activation function and model definition
We will use the `sigmoid` activation function. Please create a function that returns the output of the sigmoid, $\sigma(z)$ where $\sigma(z) = \frac{1}{1+\exp(-z)}$
```{r}
# activation function
sigmoid <- function(z){
  # TODO
  #output = 
  #return(output)
}
```
We will create a model with a two-dimensional input layer, one hidden layer, and a one-dimensional output layer. 
$$y = \sigma(W^{[2]}(\sigma(W^{[1]}x + b^{[1]}))+b^{[2]})$$
We define the next function that goes forward through the model. It takes the input `x`, and return the output of the model, as well as the intermediate layers that we will use in backpropagation. 

Please fill in the gaps below taking into account that: $z_2 = W^{[1]}x + b^{[1]}$, $a_2 = \sigma(z_2)$, $z_3 = W^{[2]}a_2 + b^{[1]}$, $a_3 = \sigma(z_3)$.
For this, take into account that in R matrix multiplication is performed using `%*%`, and the transposed of a matrix `x` is performed using `t(x)`. You will only need to tranpose the input `x` of the model, as in our data the observations are given per row, and we want to put them into columns. 
```{r eval=FALSE}
# forward model
forward_model <- function(model, x){
  # model is a list with the following 
  # x is the input
  W1 = model[['W1']]
  b1 = model[['b1']]
  W2 = model[['W2']]
  b2 = model[['b2']]
  
  # forward pass
  # hidden layer
  #z2 = #TODO
  #a2 = #TODO
  
  # output layer
  #z3 = #TODO
  #a3 = #TODO
  output = list(z2 = t(z2), a2=t(a2), z3=t(z2), a3=t(a3)) 
  
  return(output)
}
```

#### Loss function
Let $\mathcal{D} = \{(x^1, y^1), (x^2, y^2), \ldots, (x^N, y^N)\}$ be our training set, where $x^i\in \mathbb{R}^n$.
We define the loss function as $$J(\theta) = -\sum_{i=1}^{N} y^i \log(f_{\theta}(x^i)) + (1-y^i) \log(1-f_{\theta}(x^i))$$, which corresponds to the negative log-likelihood of of our dataset assuming it was sampled from a Bernouilli distribution. 

```{r}
loss_fn <- function(model, x, y){
  layers_model = forward_model(model, x)
  y_pred = layers_model[['a3']]
  loss = y * log(y_pred) + (1-y) * log(1-y_pred)
  return(sum(-loss))
}
```

#### Gradient descent step.
See for example (https://arxiv.org/pdf/1801.05894.pdf) for a general form of the gradient of the cost function for the parameters in each layer of the network.
The following function performs a gradient descent step using the whole training set $$\theta := \theta - \alpha \cdot \partial_{\theta}J(\theta)$$
```{r}
GD_step <- function(model, x, y, lr=0.001){
  W1 = model[['W1']]
  b1 = model[['b1']]
  W2 = model[['W2']]
  b2 = model[['b2']]
  
  pred_model = forward_model(model, x)
  z2 = pred_model[['z2']]
  a2 = pred_model[['a2']]
  z3 = pred_model[['z3']]
  a3 = pred_model[['a3']]
  
  delta3 = a3 - y
  dW2 = t(a2) %*% delta3
  db2 = apply(FUN=sum, X=delta3, MARGIN=2)
  
  delta2 = sigmoid(z2) * (1-sigmoid(z2)) * (delta3%*%W2)
  dW1 = t(x) %*% delta2
  db1 = apply(FUN=sum, X=delta2, MARGIN=2)
  
  W2 = W2 - lr * t(dW2)
  b2 = b2 - lr * db2
  W1 = W1 - lr * t(dW1)
  b1 = b1 - lr * db1
  
  model[['W1']] = W1
  model[['b1']] = b1
  model[['W2']] = W2
  model[['b2']] = b2
  
  return(model)
}
```

#### We put everything together for the training:
- Initialise $W^{[1]}, b^{[1]}, W^{[2]}, b^{[2]}$.
- While Not convergence:
    - Calculate $J(\theta)$
    - Update $W^{[i]} := W^{[i]} - \alpha \cdot \partial_{W^{[i]}}J$
    - Update $b^{[i]} := b^{[i]} - \alpha \cdot \partial_{b^{[i]}}J$
    
```{r eval=FALSE}
train <- function(model, n_epochs, x, y){
  for(epoch in 1:n_epochs){
    model = GD_step(model, x=x, y=y)
    loss = loss_fn(model, x=x, y=y)
    
    if(epoch %% 10 == 0){
      log = paste("Epoch: ", epoch, "/", n_epochs,", loss: ", loss, sep="")
      print(log)
    }
  }
  return(model)
}

n_hidden = 30
matrix
W1 = matrix(data=rnorm(2*n_hidden), nrow=n_hidden, ncol=2)
b1 = rnorm(n_hidden)
W2 = matrix(data=rnorm(n_hidden), nrow=1, ncol=n_hidden)
b2 = rnorm(1)
model = list(W1 = W1, b1=b1, W2=W2, b2=b2)

model = train(model=model, n_epochs=10000, x=as.matrix(x=data[,c("x1", "x2")]), y=c(x=data$y))
```

We plot the results with the new model.
```{r eval=FALSE}
grid = expand.grid(x1 = seq(min(data$x1), max(data$x1), 0.02),
                   x2 = seq(min(data$x2), max(data$x2), 0.02))
y_grid = forward_model(model, x=as.matrix(grid))[['a3']]
y_grid_binary = y_grid>0.5
grid$y = as.numeric(y_grid_binary)

p = ggplot(data=data, aes(x1,x2, color=factor(y))) + geom_point(size=1) 
p = p + geom_point(data=grid, aes(x1, x2, color=factor(y)), alpha=0.1)
print(p)
```











